---
title: "Data Table"
output:
  html_document: default
  html_notebook: default
---
# Here we will be learning all about data.table package.

The content is from datacamp course: Data analysis - The data table way

```{r,include=FALSE}
library(data.table)
```

`data.table` package is loaded

The format of a data table syntax is 
`DT[i, j, by]` which stands for 
* Take DT, subset by `i`, then calculate `j` grouped by `by` *

The SQL equivalents for i, j, by in R are given by WHERE, SELECT & GROUP BY respectively

We will cover each of the parts one by one

There are 3 things we need to know about datatypes
    
* `typeof(1)` is double or numeric but `typeof(1L)` is integer
* `typeof(NA)` is logical but `typeof(NA_integer_)` is integer
* character strings are very efficient in R due to use of global cache

## `i` part

###`i` is the subsetting part
```{r}
DT <- data.table(A = 1:5, B = c('a', 'b', 'c', 'd', 'e'), C = TRUE)
DT #The smaller vector is recycled
DT[2:3,]
DT[2:3] #Gives the same result as the previous code
```

###A data.table is also a data.frame
```{r}
class(DT)
```
Hence, the basic rules of data.frame also work on the variable DT

###`.N` stores the number of rows of a data table
```{r}
DT[.N] #Results in the last row of DT
DT[nrow(DT)] #Gives the same result
DT[.N - 2] #Third last row of DT 
```


## `j` part

### `j` is the column selection part

We can select columns using `.()` (dot-paranthesis syntax)

```{r}
DT[, .(B, C)] #Selecting columns B and C
```

### We can not only select columns, but throw pretty much everything at `j`

```{r}
DT[, .(A, B, C, loga_mynewcolumn = log(A))] #Printing a data table containing logarithmic operation column on A
DT[, .(B, C = sum(A))] # C is recycled
DT[, log(10)] #This results in a vector, not a data table
# Notice the syntax.. This doesn't contain our dot-paranthesis syntax
DT[, {print(10)
  print(15)
  print(1:5)}] #We are throwing 3 printing statements at `j`
```

### We can also use .N in `j`

This is how it works: It only calculates the number of rows in the group. When it is grouped with `by` argument, it is even more powerful (Will be shown later)

```{r}
DT <- data.table(iris)
DT[,.N]
```

## `by` part

by is the part where the data table is _grouped by_ or _summarized_

Once, again lets recall the basic syntax of a table:
`DT[i, j, by]` stands for Take `DT`, subset by `i`, then calculate `j` grouped by `by`

### Now, lets calculate the mean Sepal Length for each species

```{r}
DT[, .(mean(Sepal.Length)), by = .(Species)]
DT[, mean(Sepal.Length), by = Species] #Gives the same result..
# Notice the syntax.. we have skipped the dot-paranthesis syntax
# This can be done if there is only one argument for 'j' and 'by'
DT[, mean(Sepal.Length), Species] #The same output again
```

### Using .N in `j`

In the below code, we will use .N in `j`, which will calculate the no. of rows for each group
```{r}
DT[, .N, by = Species]
DT[, .(count = .N), by = Species] #Same output, column name changed

# Now we will use an expression in 'by'
DT[, .(count = .N), by = .(Starting_with = substr(Species,1,1))]
```

### Using expressions in `by`

Using rounded area to group the specimens
Now, lets group the specimens by Sepal Area(rounded to the nearest 10 cm2) and count how many for each group
```{r}
DT[, .N, by = .(10 * round(Sepal.Length * Sepal.Width / 10))]
DT[, .(Count = .N), by = .(Area = 10 * round(Sepal.Length * Sepal.Width / 10))] # Naming the columns
```


## Chaining operations

Chaining operations comes very handy, when we have multiple operations to do with a datatable. It helps to avoid those pesky objects that can quickly clutter up the workspace

```{r}
DT[, .(
  Sepal.Length = mean(Sepal.Length),
  Sepal.Width = mean(Sepal.Width),
  Petal.Width = mean(Petal.Width),
  Petal.Length = mean(Petal.Length)
  ), by = Species][order(Species)]
# This calculates the mean of each of the columns, grouped by Species
```

Lets try it an easier way
```{r}
DT[, lapply(.SD, mean), by = Species]
```
`.SD` stores the subset of data for each unique value of `by` argument
So, the above command applies mean for all the _subset_ groups of the data table, grouped by Species.

Similar to `.SD`, we have `.SDcols` which can be used to subset columns
Lets say, we want to perform the same `mean` of only the Sepal.Length column
```{r}
DT[, lapply(.SD,mean), by = Species, .SDcols = "Sepal.Length"]
# OR
DT[, lapply(.SD, mean), by = Species, .SDcols = 1]
# Apply mean for those columns starting with Petal
DT[, lapply(.SD, mean), by = Species, .SDcols = grep("Petal", colnames(DT))]
```

As we can see, there is no `.()` around lapply, because it returns a list

## Next we will learn abt the following things:
1. := operator, `:=` functional operator, removing and updating columns by reference.
2. set functions - set, setnames, setcolorder
3. indexing
4. setkey

### `:=` operator

`:=` lets us assign/modify/ create columns by reference. In such a way, we dont need to give the assignment operator

Lets try and delete columns from a data table and see

```{r}
DT[,c("Sepal.Length", "Sepal.Width") := NULL] # Deleting 2 columns
DT[,Petal.Width := NULL] # Shorthand for deleting a column
DT
```
`:=` can also be used to set column values
```{r}
DT[, Species := c("a", "b", "c")] # Recycling vector ("a","b","c") in the 2nd column
DT
DT[, c("Petal.Length", "Species") := list(1:150, 51:200)] # Modifying 2 columns at once
DT
```

#### `:=` Functional operator
``:=`` can also be used to modify multiple columns
```{r}
DT[, `:=`(Petal.Length = NULL, Species = 61:210)]
DT
```

Notice the back-ticks (``) in the operator

One of the most important things of understanding the concepts of data table is that if a single variable is defined outside of a data table, it is evaluated in the outside environment and it is evaluated inside the scope of the data table, if otherwise

```{r, include=FALSE}
DT <- data.table(iris)
```

Another way of modifying columns and elements of a data table is by set commands

### Set functions

These include `set`, `setnames`, `setcolorder`

Lets say, we want to set `NA` values for 3 random rows for each of the columns `Sepal.Width` and `Sepal.Length`
```{r}
for (j in 1:2) {
  set(DT, i = sample(150, 3), j = j, NA)
}
rm(j)
DT[is.na(Sepal.Length)] # Indexing of rows (in the `i` part)
DT[Sepal.Length>7]
```

`setnames()` is for setting column names and `setcolorder()` is for setting column orders
```{r}
# Changing column names starting from Sepal. to "".. Removing the "Sepal." prefix
setnames(DT, old = colnames(DT), new = gsub("Sepal.","",colnames(DT)))
DT #Now look at our data table
setcolorder(DT, neworder = c("Species", "Length", "Width", "Petal.Length", "Petal.Width"))
DT
```

### Keys

Setkey is very powerful function for data tables. It is kinda like a power optimizer for data tables
```{r}
setkey(DT, "Species")
DT
```

It sorts the row names and it provides many other functionalities too

```{r}
set.seed(100)
x <- data.table(letter = sample(letters[1:10],50, replace = TRUE), number = sample(10, 50, replace = TRUE), logical = c(TRUE,FALSE))
x # Before setting keys
setkey(x, "letter"); x #After setting "letter" column as key
x["a"]
```

By setting a "key" column, we allow it to be used as super-charged row names. We can also setup 2 column keys
```{r}
setkey(x,letter, number)
x # Take a look at it.. The ordering has changed
x[c("a","f")] # Looking at two groups at once
# By default, all the rows are returned. This behaviour can be changed using mult argument
x[c("a","f"), mult = "first"]
x[c("b","c")]
x[c("b","c"), mult = "last"]
x[.("b",7)] # Indexing 2 columns, letter and number.. Notice the .() syntax again
x[.("c",10)]
x[c("b","c","z")] # NA introduced for no possible match
x[c("b", "c", "z"), nomatch = 0] #NA value removed
x[.(c("b","c"),c(3,10))]
x[.(c("b","c"),c(10,3))] #NAs introduced
x[.(c("b","c","d"),c(3,10,11))]
x[.(c("b","c","d"),c(3,10,11)), nomatch = 0] # Skipped NAs
```

Now, we will use the `.EACHI` argument to evaluate each of the `i` groups
```{r}
x[c("b","c")]
x[c("b","c"), .SD[c(1,.N)], by = .EACHI]
# In the above step, we returned the first and last rows of each of the groups
# In the next step, we will return the groups before returning the first and last rows of the groups
x[c("b","c"), { print(.SD); .SD[c(1,.N)] }, by = .EACHI]
x[c("b", "c"), c(print(.SD) , .SD[c(1, .N)]), by = .EACHI] # The same result
```

So as we can see, we can throw pretty much anything and everything at `j`, as long as it returns a list
```{r}
x[c("b","c"), summary(x), by = .EACHI]
```

We can extract the keys of a data table by using `keys()` function
```{r}
key(x)
```

## Rolling joins

In the previous section, we learnt about equijoins, the fact that we could find the contents of a data table that match a key index, for e.g. "b" and "c" in the previous section

In this section, we will learn abt rolling joins, finding contents of a data table that do not match the indices provided
```{r}
x[,newnu:= sample(10,replace = TRUE)] #Adding a new column
x[,logical:=NULL] #Removing logical column
x <- x[1:10] #Limiting ourselves to 10 rows
x
x[.("b")]
x[.("b",3)] #Equijoin
x[.("b",4)] # Returns NA
x[.("b",4), roll = TRUE] #Rolling to the prevalent row
x[.("b",6),roll = TRUE]
x[.("b",6), roll = "nearest"] #Rolling to nearest row
x[.("b",0:10)] #Contains NA values for non-matching rows
x[.("b",0:10), roll = +Inf] #Forward roll
x[.("b", 0:10), roll = +Inf, rollends = TRUE] #The values before the first match.. for number values 0,1,2 are assigned values
x[.("b",0:10),roll = -Inf] #Backward roll
x[.("b",6),roll = TRUE]
x["b"]; x[.("b",6),roll = 2] #Searching for a matching row within 2 row window
x["b"]; x[.("b",6),roll = 3] #Successfully searching for a matching row within 3 previous rows
x["b"]; x[.("b",4),roll = -Inf]
x["b"]; x[.("b",4),roll = -2] # Trying to match backwards for a matching row
x["b"]; x[.("b",4),roll = -3] # Successfully matched
```

Thoroughly enjoyed learning Data table. I will be using it prominently in my analyses
