---
title: "Data Table"
output:
  html_document: default
  html_notebook: default
---
# Here we will be learning all about data.table package.

The content is from datacamp course: Data analysis - The data table way

```{r,include=FALSE}
library(data.table)
```

`data.table` package is loaded

The format of a data table syntax is 
`DT[i, j, by]` which stands for 
* Take DT, subset by `i`, then calculate `j` grouped by `by` *

The SQL equivalents for i, j, by in R are given by WHERE, SELECT & GROUP BY respectively

We will cover each of the parts one by one

There are 3 things we need to know about datatypes
    
* `typeof(1)` is double or numeric but `typeof(1L)` is integer
* `typeof(NA)` is logical but `typeof(NA_integer_)` is integer
* character strings are very efficient in R due to use of global cache

## `i` part

###`i` is the subsetting part
```{r}
DT <- data.table(A = 1:5, B = c('a', 'b', 'c', 'd', 'e'), C = TRUE)
DT #The smaller vector is recycled
DT[2:3,]
DT[2:3] #Gives the same result as the previous code
```

###A data.table is also a data.frame
```{r}
class(DT)
```
Hence, the basic rules of data.frame also work on the variable DT

###`.N` stores the number of rows of a data table
```{r}
DT[.N] #Results in the last row of DT
DT[nrow(DT)] #Gives the same result
DT[.N-2] #Third last row of DT 
```


## `j` part

### `j` is the column selection part

We can select columns using `.()` (dot-paranthesis syntax)

```{r}
DT[, .(B, C)] #Selecting columns B and C
```

### We can not only select columns, but throw pretty much everything at `j`

```{r}
DT[, .(A, B, C, loga_mynewcolumn = log(A))] #Printing a data table containing logarithmic operation column on A
DT[, .(B, C = sum(A))] # C is recycled
DT[, log(10)] #This results in a vector, not a data table
# Notice the syntax.. This doesn't contain our dot-paranthesis syntax
DT[, {print(10)
  print(15)
  print(1:5)}] #We are throwing 3 printing statements at `j`
```

